The promise of community driven preprints in ecology and evolution

### 2024-04-29 ms/ms.qmd
Fixing up comments from Losia and Shinichi. Need to work on figures, conclusion and reporting a few numbers..just check

Authors: Daniel W.A. Noble ^1^ $\ddagger$, Shinichi Nakagawa^2^, Hackathon Participants, Malgorzata Lagisz^2^

@@

@@

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While preprint use in biology has had a slower uptake [@Berg2016], in ecology and evolution, new discipline specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Such practices are expected to promote open science and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally.

Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings; 2) encouraging constructive and critical feedback from a research community; 3) helping establish precedence of research findings; and 4) removing financial barriers to publication [@Proulx2013; @bourne2017ten; @vale2015accelerating]. *EcoEvoRxiv* is a community driven preprint server that has paved the way for new initiatives, including accepting multilingual preprints, recognising and accepting registered reports and non-traditional research reports, and promoting peer review and community discussion to improve the quality of preprints and speed up their peer-reviewed publication (e.g., Peer Community In -- PCI).

Preprint servers, such as *EcoEvoRxiv*, can empower researchers to make their findings more accessible, open and transparent but only if they are used well and initiatives are widely known. Significant barriers to preprint use still exist globally – from lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by a research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution which can help inform future initiatives, improve others, and allow us to work harder in further break down barriers to their use in our field.

Here, we explore the first preprints uploaded to *EcoEvoRxiv*, one of the few community run preprint servers, to understand preprint practices and use in ecology and evolution. We focus our attention on understanding better: 1) what countries make use of preprints; 2) the extent to which authors make use of preprint servers for registered reports and community driven peer review  3) how career stage and gender impact preprint use; 4) whether data and code are more likely to be shared in preprints; and 5) how long it takes for a preprint to become published, and how many preprints remain unpublished. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other servers which we think will help further educate researchers about the benefits of using community driven preprint servers to post their research findings.

@@

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolution (SORTEE). Editors are researchers across the globe in the fields of ecology and evolution who volunteer their time to screen papers and push new initiatives in the preprint space. To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the articles available on *EcoEvoRxiv* as of `r gsub("UTC", "", dates[2])`.  *EcoEvoRxiv* accepts both preprints and postprints. Postprints are articles that are already published but are the final version of the published paper before typesetting for the research journal. Postprints are a useful way to ensure published articles are openly accessible and can be posted anytime after publication (assuming the journal has a preprint policy that recognises preprints - though most do). For the purpose of summarising articles on *EcoEvoRxiv* we include both preprints and postprints in our 'demographics' analysis.

```{r, fig-summary}
#| label: fig-summary
#| fig-cap: Summary of articles posted to *EcoEvoRxiv*. A) Number of preprints and postprints published monthly to *EcoEvoRxiv* between 2018 and 2023; B) Countries that are posting articles (preprints and postprints) to *EcoEvoRxiv*; C) Taxa used / covered in the articles posted to *EcoEvoRxiv*; D) Types of articles accepted on *EcoEvoRxiv*. E) Career stage of authors posting to *EcoEvoRxiv* along with the gender of the submitting author.
#| echo: false
#| warning: false
#| message: false

@@

@@

### *Demographics of **EcoEvoRxiv** preprints (and postprints)*

After removing `r dim(dupl)[1]` duplicate titles – suggesting a few authors created multiple submissons for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55-60 preprints published each month in the last two years ([@fig-summary]A). A total of `r length(unique(countries$country))-1` countries have used *EcoEvoRxiv* as a preprint server. Unsurprisingly, countries in North America and Europe preprint more exstensively. Countries in Africa, Central America and parts of Asia are still slow to use *EcoEvoRxiv* ([@fig-summary]B). Overall, 90% of the articles come from just `r nrow(most_use)-1` countries. Articles covered all major taxomic groups, with the most common being vertebrates (`r sum_taxa$perc[6]`%), plants (`r sum_taxa$perc[1]`%) and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]C).

@@

Given that *EcoEvoRxiv* is a community driven preprint server it has managed to diversify the types of articles accepted. It accepts reports and books that are often not easily accessible, but may contain important data of relevance to research syntheses. Accepting a greater diversity of article types allows *EcoEvoRxiv* to begin dealing with the 'grey literature' problem [@haddaway2015shades; @haddaway2020eight]. As a result, articles on *EcoEvoRxiv* are more diverse in nature than other preprint servers; nonetheless, typical research articles (`r sum_types[sum_types$type == "research article",]$perc`%) are still the most common type of preprint, followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) ([@fig-summary]D).

Reports, parcticularly from government and industry research areas exist but are not yet common on *EcoEvoRxiv*. We would encourage authors to consider posting these types of articles to ensure they are openly accessible and more easily found. *EcoEvoRxiv* has already formed fruitful partnerships with the International Union for Conservation of Nature (IUCN) to be a place where their reports can be lodged to make them more easily accessible to the broader research community. For example, IUCN Red Ecosystem Reports are now posted to *EcoEvoRxiv* and our community has been able to work closely with the IUCN to ensure these reports meet the IUCN requirements. Such reports often contain important data that can be leveraged for research syntheses and meta-analyses.

@@

A significant barrier to the communication of research findings is due to the fact that they are primarily communicated in English [@amano2016languages; @amano2013four; @amano2021ten]. Communication of research through a single language has major consequences for the global distribution of knowledge resulting in knowledge gaps in some of the most biodiverse and threatened regions in the world [@amano2023role; @zenni2023multi]. Such gaps also have flow on effects for research syntheses and meta-analyses because it creates a distorted picture of our knowledge base that can lead to bias and inadequate sampling that affects policy development and decision making [@hannah2024language; @zenni2023multi; @white2021geographical; @konno2020ignoring].

*EcoEvoRxiv* is the only preprint server breaking down language barriers to scientific communication; currently accepting English, Spanish, Portuguese and French manuscripts. Such initiatives are incredibly important if we are to begin filling global voids of scientitic knowledge. Multilingual initiatives have been slow to take off on *EcoEvoRxiv*, with only a few Spanish preprints posted to date, and none in other accepted langauges (e.g., French and Portuguese). Part of the challenge in getting authors to submit non-English preprints is the lack of awareness of *EcoEvoRxiv* in non-English speaking countries and the fact that it accepts preprints in different languages, cultural differences in the perception of preprints and a strong reliance on traditional publishing models, which typically favour publishing in English.

@@

Research papers take time to be published (see below). Early and mid career researchers (ECMRs) need their research to come out quickly for job applications, promotion and obtaining grants to accerate their careers [@vale2015accelerating; @sarabipour2019value]. Preprints are one way ECMRs can acheive faster dissemination and greater visibility. As such, we expect ECMRs to accept and adopt preprints more than later career colleagues. At the same time, gender differences in preprint use and publication outcomes  have been established in a number of other research fields, including in ecology and evolution [@fox2019gender; @wehner2020comparison]. We therefore expect such discrepencies to manifest in preprint use on *EcoEvoRxiv*, but it is unclear to what extent. Understanding gender publishing patterns is challenging with observational data such as ours because we cannot know the gender of authors for certainty but we can use a data driven approach to ascertain the probability that a particular name is of a given gender (male or female). To obtain an idea of an authors' gender, we used the R package *DemografixeR* (vers. `r utils::packageVersion("DemografixeR")`; @RJ-2016-002) to predict the most likely gender of the submitting author of a preprint. We incorperated information about the submitting authors country of affiliation to better account for differences across countries. We only used names where the gender was identified with 95% certainty.

Overall, females were less likely to publish preprints on *EcoEvoRxiv* compared to males (females: `r (table(data4$gender)[1] / sum(table(data4$gender))*100)`% compared to males: `r (table(data4$gender)[2] / sum(table(data4$gender))*100)`%), which is likely a reflection of the broader publishing disparities between male and female scientists [@fox2019gender]. There was also strong evidence that the number of preprints posted decreases with later career stages (negative binomial glm: year slope = `r res[2, "Estimate"]`, SE: `r res[2, "Std. Error"]`, *p* < 0.001, N = `r nrow(count_dat[complete.cases(count_dat),])`). The majority of preprints were submitted by authors who published their first paper in the last 10 years ([@fig-summary]E), with the median year of first publication of submitting authors being `r median(data2$submitting_author_first_publication_year, na.rm = TRUE)` (mean = `r round(mean(data2$submitting_author_first_publication_year, na.rm = TRUE),0)`; SD = `r sd(data2$submitting_author_first_publication_year, na.rm = TRUE)`). These preprinting practices support the idea that ECMRs may value preprints as a way to make their work more visible and disseminate their finding more quickly.

@@

### *Science takes time, but publication could take longer*
Increased competition in science has resulted in the bar being raised with respect to the amount of data required for publication [@vale2015accelerating]. This is a good outcome if it results in higher impact research that better clarifies our understanding of the natural world, but it does come at a cost for the rate at which research findings can be disseminated [@vale2015accelerating]. Preprinting has been proposed as a way research can be disseminated more quickly as it can take a long time before results are ultimately published [@bourne2017ten; @vale2015accelerating]. However, data on the time to publication is sparse, and the real benefit of preprints in this context would be reinforced if preprinting is able to dramatically speed up dissemination. We evaluated just how much time it takes to publish a research paper in ecology and evolution by ascertaining the time between when a preprint was first posted to when it was finally accepted in a research journal. In total, `r unpub` papers were still unpublished out of the `r nrow(data2)` (`r (unpub / nrow(data2))*100`%) at the time these data were collected. Astonishingly, the maximum time between submission and publication was `r sum_preprints$max` days ([@fig-pubsummary]A), and the median time to publication was `r sum_preprints$median` days (mean = `r sum_preprints$mean`; SD = `r sum_preprints$sd` days). Our results largely confirm the long timespans most authors are dealing with to get their research published after the paper is written.

### *Cautious 'open'-mindedness of research in preprints*
In additon to speeding up dissemination, preprints can also be a useful way to ensure that research remains open and accessible to the research community because the paper is available online no matter what the open access status of the publication ends up being [@bourne2017ten; @vale2015accelerating]. Using the published articles on *EcoEvoRxiv*, we evaluated the open access status of each paper using....LOSIA?. Most of the published preprints ended up being open access([@fig-pubsummary]A), however, the status of many (i.e., `r sum_is_oa$n[3]` or `r sum_is_oa$percent[3]*100`%) were unknown, and `r sum_is_oa$n[1]` or `r sum_is_oa$percent[1]*100`% were not published in open access reserach journals. The level of open access for a paper published in an open access journal also varied widely ([@fig-pubsummary]B).

Data and code sharing is also a key component of open science [@roche2015public]. In the spirit of 'openness' we expected data and code sharing among preprints to be better than many papers published in research journals [@roche2015public; @o2021preferred]. Despite this, we found that `r preprint_data[1,2]` (`r preprint_data[1,"percent"]*100`%) of relevant preprints on *EcoEvoRxiv* (i.e., only where data was generated) did not share data, while `r preprint_code[1,2]` (`r preprint_code[1,"percent"]*100`%) did not share code. Authors may be reluctant to share data in code for preprints because of the perceived concern that others may acquire and use such data and code without authors' consent. However, while `r pre_article_data[3,2]` (`r pre_article_data[3,"percent"]*100`%) of preprints that did not share data at the preprint stage did ultimately share data in the published article, there were still `r pre_article_data[2,2]` (`r pre_article_data[2,"percent"]*100`%) that never ended up sharing data. The same was true for code. Overall, `r pre_article_code[3,2]` (`r pre_article_code[3,"percent"]*100`%) of preprints did not share code at the preprint stage but did in the published article, however, there were still `r pre_article_code[2,2]` (`r pre_article_code[2,"percent"]*100`%) that never ended up sharing code. Code and data sharing patterns in our sample of preprints is consistent with other analyses looking at code and data sharing in published articles (e.g., [@o2021preferred]), even for journals that have strict public data archiving policies [@roche2015public].

# Paving *our* own future of open, transparent and community driven science

Our data driven approach has allowed us to better understand some of the key elements of uncertainty surrounding preprints and identify shortcomings and areas needing greater clarity to authors. A number of key insights can be gleaned upon from an analysis of the first *EcoEvoRxiv* preprints.

- First, authors need not be afraid of sharing data and code early in the preprint stage. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings [@sarabipour2019value]. If authors are worried about data being used unintentially, clear information surrounding its re-use and withholding some data or meta-data needed to understand the dataset may be options. 
- Second, authors need to be aware of the benefits of posting preprints and postprints for constructive feedback [@bourne2017ten]. While it is clear preprints help establish precedence and allow findings to be openly accessible it is still rare that constructive discussions form around preprints in an open form, which is consistent with findings on other preprint servers (e.g. BioRxiv [@anderson2020biorxiv]). It is clear that we, as a community, need to provide better platforms that document discussions around a preprint findings. Such discussions are useful for the authors to help improve their work and communicate their findings more effectively (when done constructively, of course). One way to facilitate such discussions may be to use open peer reviewing serves such as *Peer Community In* (PCI) or *PubPeer* to provide feedback on preprints that ultimately are associated with a preprint.
- Third, it remains clear that the time between preprinting and publication is still quite long. One possible explanation is that preprints are not being sent to journals deemed suitable for the work. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recomendations when deciding to publish a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process. 
- Finally, while most preprints get seemlessly connected with and merged with their published version some remain disconnected. Incorrect cross-linking can create confusion and lead to frustration among authors. Thankfully, the reasons for unmatched preprints and publications are fairly well understood and easily rectified. They often result from a mismatch between titles or authors in preprint and published metadata. In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%). Mismatched metadata was almost always the main contributer to preprints and published articles not being matched automatically in Google Scholar during our data collection. We recommend that authors make sure that, when preprints end up getting published, they update their preprint with the publication DOI when it is available. Particularly if their title has changed. This is quite easy for authors to do in *EcoEvoRxiv* and it will ensure that the preprint is linked to the published article and the citations are merged. It will also create a link between the final preprint and the published paper so that, no matter the open access status of the publication, the reserach findings will remain openly accessible, saving costs for authors and improving the visibility and use of research.

Despite the early successes of the new initiatives taken by *EcoEvoRxiv* it is still clear that much work needs to be done to improve our communities understanding of preprints (and postprints). We view this perspective piece as one small step in achieving that goal. Authors also need to be aware of the benefits of using community driven preprint servers and the unique initiatives being pursued. Community driven preprint servers are capable of setting their own agenda and are driven by the needs and disires of the community they would like their findings to be shared with. Supporting these initiatives should be a priority for all researchers and those volunteering their time at *EcoEvoRxiv* remain open to new and innovative ways to improve publication and open science practices.  We believe that the future of preprints is bright and community driven initiatives, such as *EcoEvoRxiv*, are set to play a key role in the future of scientific publishing.

# Awknowledgements
We would like to thank the California Digitial Library (CDL) and the CDL team (particularly Alainna Wrigley) for their support in hosting and maintaining *EcoEvoRxiv* for the Society for Open, Reliable, and Transparent Ecology and Evolution (SORTEE). We would also like to thank Gabriela Hidalgo and Daisy Larios for helping connect us with the IUCN and facilitating discussions to make *EcoEvoRxiv* a place where IUCN reports can be posted. Finally, we would also like to thank the endless number of SORTEE volunteers, and those especially on the *EcoEvoRxiv* Commitee, who have helped to make *EcoEvoRxiv* a success. This paper emerged from a hackathon at the 2022 SORTEE conference, and we thank delegates who attended the session but could not be part of this paper. DWAN would also like to thank the Australian Research Council for a Future Fellowship (FT220100276).

>>>>

Authors: Daniel W.A. Noble ^1^, $\ddagger$, Hackathon Participants, Shinichi Nakagawa^2,\*^,Malgorzata Lagisz^2,\*^

@@

\* equal contribution

@@

Publishing preprints -- papers communicating non-peer-reviewed research findings -- is now an entrenched practice across a multitude of scientific disciplines [@Ginsparg2011]. While preprint use in biology has had a slower uptake [@Berg2016], in ecology and evolution, new discipline-specific preprint servers, such as *EcoEvoRxiv*, provide a means by which ecologists and evolutionary biologists can disseminate research findings. Preprints are expected to promote open science and facilitate the rapid sharing of new discoveries that can have significant impacts on fundamental and applied knowledge globally. Preprints attempt to break down barriers to scientific publishing by: 1) increasing the speed, visibility and output of research findings; 2) helping establish precedence of research findings; 3) removing financial barriers to publication; and 4) encouraging constructive and critical feedback from a research community [@Proulx2013; @bourne2017ten; @vale2015accelerating].

Preprint servers can empower researchers to make their findings more accessible, open and transparent but only if they are used well and initiatives are widely known. Significant barriers to preprint use still exist globally – from lack of clarity around preprint policies at journals [@Klebel2020] to a stigma that preprints are viewed by a research community to be of poor quality [@Chiarelli2019] (but see [@Faser2022]). Nonetheless, we lack a systematic understanding of the factors that influence preprint use in ecology and evolution which can help inform future initiatives, improve others, and allow us to work harder in further breaking down barriers to their use in our field.

*EcoEvoRxiv* is one of the few community driven preprint servers that has paved the way for new initiatives, including accepting multilingual preprints, recognising and accepting registered reports and non-traditional research reports, and promoting peer review and community discussion to improve the quality of preprints and speed up their peer-reviewed publication (e.g., Peer Community In -- PCI).

Here, we explore the first preprints uploaded to *EcoEvoRxiv* to understand preprint practices and use in ecology and evolution. We focus our attention on understanding better: 1) what countries authors that preprint are located; 2) how career stage and gender impact preprint use; 3) the extent to which authors make use of preprint servers for registered reports and community-driven peer review; 4) whether data and code are more likely to be shared in preprints; and 5) how long it takes for a preprint to become published, and how many preprints remain unpublished. In the process, we also provide a summary of what makes *EcoEvoRxiv* distinct from other servers which we think will help further educate researchers about the benefits of using community driven preprint servers to post their research findings.

@@

*EcoEvoRxiv* is run by the Society for Open, Reliable, and Transparent Ecology and Evolution (SORTEE). Originally launched in 2018 on the Center for Open Science preprint platforms, *EcoEvoRxiv* has grown to become a popular preprint server for researchers in ecology and evolution. It has since been adopted by the California Digital Library (CDL) to be hosted in perpetuity without fees. Editors are researchers across the globe in the fields of ecology and evolution who volunteer their time to screen papers and push new initiatives in the preprint space. *EcoEvoRxiv* accepts both preprints and postprints. Postprints are articles that are already published but are the final version of the published paper before typesetting for the research journal. Postprints are a useful way to ensure published articles are openly accessible and can be posted anytime after publication (assuming the journal has a preprint policy that recognises preprints - though most do). To better understand preprint (and postprint) use on *EcoEvoRxiv*, we downloaded metadata on the articles available on *EcoEvoRxiv* as of `r gsub("UTC", "", dates[2])`. For the purpose of summarising articles on *EcoEvoRxiv* we include both preprints and postprints as 'articles' in our overview analysis.

```{r, fig-summary}
#| label: fig-summary
#| fig-cap: Summary of articles posted to *EcoEvoRxiv*. A) Number of articles (preprints and postprints) published monthly to *EcoEvoRxiv* between 2018 and 2023; B) Countries that are posting articles (preprints and postprints) to *EcoEvoRxiv* as inferred from the country of affiliation of the submitting author; C) Taxa used / covered in the articles posted to *EcoEvoRxiv*; D) Types of articles accepted on *EcoEvoRxiv*. E) Career stage of authors posting to *EcoEvoRxiv* along with the gender of the submitting author.
#| echo: false
#| warning: false
#| message: false

@@

nt <- sum_types  %>% filter(type %in% c("book", "report", "book chapter", "other"))  %>% summarise(perc = sum(perc))

@@

### *Overview of **EcoEvoRxiv** preprints (and postprints)*

After removing `r dim(dupl)[1]` duplicate titles – suggesting a few authors created multiple submissions for the same preprint rather than updating an existing submission – we had data for a total of `r pr` articles with ~55-60 preprints published each month in the last two years ([@fig-summary]A). Authors from a total of `r length(unique(countries$country))-1` countries have used *EcoEvoRxiv* as a preprint server. Unsurprisingly, authors from countries in North America and Europe upload the most preprints. Authors from countries in Africa, Central America and parts of Asia still do not use *EcoEvoRxiv* much ([@fig-summary]B). Overall, 90% of the articles come from just `r nrow(most_use)-1` countries. Articles covered all major taxonomic groups, with the most common being vertebrates (`r sum_taxa$perc[6]`%), plants (`r sum_taxa$perc[1]`%) and invertebrates (`r sum_taxa$perc[5]`%) ([@fig-summary]C).

@@

Given that *EcoEvoRxiv* is a community-driven preprint server it has managed to diversify the types of articles accepted. It accepts reports and books that are often not easily accessible, but may contain important data of relevance to research syntheses or for practitioners.  As a result, articles on *EcoEvoRxiv* are more diverse in nature than those on other preprint servers; nonetheless, typical research articles (`r sum_types[sum_types$type == "research article",]$perc`%) are still the most common type of preprint, followed by reviews and meta-analyses (`r sum_types[sum_types$type == "reviews and meta-analyses",]$perc`%) and opinion papers (`r sum_types[sum_types$type == "opinion",]$perc`%) ([@fig-summary]D).

Accepting a greater diversity of article types allows *EcoEvoRxiv* to begin dealing with the 'grey literature' problem whereby data that are relevant for research syntheses are not published in typical peer-reviewed journals[@haddaway2015shades; @haddaway2020eight]. This is reflected by `r nt`% of articles that could be considered ‘grey literature’ in ecology and evolutionary biology: books, book chapters, reports, and other research output types. Currently, *EcoEvoRxiv* does not host many reports (`r sum_types[sum_types$type == "report", "perc"]`%), particularly from government and industry. We would encourage authors to consider posting these types of articles to ensure they are openly accessible and more easily found. *EcoEvoRxiv* has already formed fruitful partnerships with the International Union for Conservation of Nature (IUCN) to be a place where their reports can be lodged to make them more easily accessible to the broader research community. For example, IUCN Red Ecosystem Reports are now posted to *EcoEvoRxiv* and our community has been able to work closely with the IUCN to ensure these documents meet the IUCN requirements.

@@

A significant barrier to the communication of research findings is due to the fact that they are primarily communicated in English [@amano2016languages; @amano2013four; @amano2021ten]. Communication of research through a single language has major consequences for the global distribution of knowledge resulting in knowledge gaps in some of the most biodiverse and threatened regions in the world [@amano2023role; @zenni2023multi]. Such gaps also have flow on effects for research syntheses and meta-analyses because they create a distorted picture of our knowledge base that can lead to bias and inadequate sampling that affects policy development and decision making [@hannah2024language; @zenni2023multi; @white2021geographical; @konno2020ignoring].

*EcoEvoRxiv* is the only preprint server breaking down language barriers to scientific communication; currently accepting English, Spanish, Portuguese and French manuscripts. Such initiatives are incredibly important if we are to begin filling global voids of scientific knowledge. Multilingual initiatives have been slow to take off on *EcoEvoRxiv*, with only a few Spanish preprints posted to date, and none in other accepted languages (e.g., French and Portuguese). Part of the challenge in getting authors to submit non-English preprints is the lack of awareness of *EcoEvoRxiv* in non-English speaking countries and the fact that it accepts preprints in different languages, cultural differences in the perception of preprints and a strong reliance on traditional publishing models, which typically favour publishing in English.

@@

Research papers take time to be published (see below). Early and mid-career researchers (ECMRs) need their research to come out quickly for job applications, promotion and obtaining grants to accelerate their careers [@vale2015accelerating; @sarabipour2019value]. Preprints are one way ECMRs can achieve faster dissemination and greater visibility. As such, we expect ECMRs to accept and adopt preprints more than colleagues at later career stages. At the same time, gender differences in preprint use and publication outcomes  have been observed in a number of other research fields, including ecology and evolution [@fox2019gender; @wehner2020comparison]. We therefore expect such discrepancies to manifest in preprint use on *EcoEvoRxiv*, but it is unclear to what extent. Understanding gender publishing patterns is challenging with observational data such as ours because we cannot know the gender of authors for certainty but we can use a data-driven approach to ascertain the probability that a particular name is of a given gender (male or female). To obtain an idea of an author's gender, we used the R package *gender* (vers. `r utils::packageVersion("gender")`; @mullen2021predict) to predict the most likely gender of the submitting author of a preprint. We only used algorithm-assigned names where the gender was identified with 95% certainty. For the remaining names, we performed manual searches to determine gender based on the pronouns and photographs from professional and personal websites (we acknowledge that our approach does not capture self-assigned and non-binary genders). Our data on gender has only two missing values – one where the first name of the submitting author was missing and the other one for a collective submission.

Overall, females were less likely to publish preprints on *EcoEvoRxiv* compared to males (females: `r (table(data4$gender)[1] / sum(table(data4$gender))*100)`% compared to males: `r (table(data4$gender)[2] / sum(table(data4$gender))*100)`%), which is likely a reflection of the broader publishing disparities between male and female scientists [@fox2019gender]. There was also strong evidence that the number of preprints posted decreases with later career stages (negative binomial glm: year slope = `r res[2, "Estimate"]`, SE: `r res[2, "Std. Error"]`, *p* < 0.001, N = `r nrow(count_dat[complete.cases(count_dat),])`). The majority of preprints were submitted by authors who published their first paper in the last 10 years ([@fig-summary]E), with the median year of first publication of submitting authors being `r median(data2$submitting_author_first_publication_year, na.rm = TRUE)` (mean = `r round(mean(data2$submitting_author_first_publication_year, na.rm = TRUE),0)`; SD = `r sd(data2$submitting_author_first_publication_year, na.rm = TRUE)`). These patterns support the expectation that ECMRs may value preprints as a way to make their work more visible and disseminate their finding more quickly.

@@

### *Science takes time, but publication could take longer*
Increased competition in science has raised the bar with respect to the amount of data required for publication [@vale2015accelerating]. This is a good outcome if it results in higher impact research that better clarifies our understanding of the natural world, but it does come at a cost for the rate at which research findings can be disseminated [@vale2015accelerating]. Preprinting has been proposed as a way to disseminate research more quickly as it can take a long time before results are ultimately published [@bourne2017ten; @vale2015accelerating]. However, data on the time to publication is sparse, and the real benefit of preprints in this context would be reinforced if preprinting can dramatically speed up dissemination. We evaluated how long it takes to publish a research paper in ecology and evolution by ascertaining the time between when a preprint was first posted and finally accepted in a research journal. In total, `r unpub` papers were still unpublished out of the `r nrow(data2)` (`r (unpub / nrow(data2))*100`%) when these data were collected. Astonishingly, the maximum time between preprint submission and publication was `r sum_preprints$max` days or `r sum_preprints$max/365` years ([@fig-pubsummary]A), and the median time to publication was `r sum_preprints$median` days (mean = `r sum_preprints$mean`; SD = `r sum_preprints$sd` days). Our results largely confirm the long timespans most authors are dealing with to get their research published after the paper is written.

### *Cautious 'open'-mindedness of research in preprints*
In addition to speeding up dissemination, preprints can also be a useful way to ensure that research remains open and accessible to the research community because the paper is available online no matter what the open access status of the publication ends up being [@bourne2017ten; @vale2015accelerating]. Using the published articles on *EcoEvoRxiv*, we evaluated the open access status of each paper using the R package *roadoi* to connect to Unpaywall platform [@jahn2024]. Most of the published preprints ended up being open access ([@fig-pubsummary]A), however, the status of many `r sum_is_oa$n[3]` (`r sum_is_oa$percent[3]*100`%) were unknown, and `r sum_is_oa$n[1]` (`r sum_is_oa$percent[1]*100`%) were not published in open access research journals. The level of open access for a paper published in an open-access journal also varied widely ([@fig-pubsummary]B).

Data and code sharing are also key components of open science [@roche2015public]. In the spirit of 'openness' we expected data and code sharing among preprints to be better than in many papers published in research journals [@roche2015public; @o2021preferred]. Despite this, we found that `r preprint_data[1,2]` (`r preprint_data[1,"percent"]*100`%) of relevant preprints on *EcoEvoRxiv* (i.e., only where data was generated) did not share data, while `r preprint_code[1,2]` (`r preprint_code[1,"percent"]*100`%) did not share code. Authors may be reluctant to share data and code for preprints because of the perceived concern that others may acquire and use such data and code without authors' consent. However, while `r pre_article_data[3,2]` (`r pre_article_data[3,"percent"]*100`%) of preprints that did not share data at the preprint stage did ultimately share data in the published article, there were still `r pre_article_data[2,2]` (`r pre_article_data[2,"percent"]*100`%) that never shared data. The same was true for code. Overall, `r pre_article_code[3,2]` (`r pre_article_code[3,"percent"]*100`%) of preprints did not share code at the preprint stage but did in the published article; however, `r pre_article_code[2,2]` (`r pre_article_code[2,"percent"]*100`%) never shared code. Code and data- sharing patterns in our sample of preprints are consistent with analyses of published articles (e.g., [@o2021preferred]), even for journals with strict public data archiving policies [@roche2015public].

# Paving *our* own future of open, transparent and community-driven science

Our data-driven approach has allowed us to better understand some of the key elements of uncertainty surrounding preprints and identify shortcomings and areas needing greater clarity to authors. Several key insights follow from our analysis of the first *EcoEvoRxiv* preprints.

- First, authors need not be afraid of sharing data and code early in the preprint stage. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings [@sarabipour2019value]. If authors are worried about data being used unintentionally, clear information surrounding its reuse and withholding some data or meta-data needed to understand the dataset may be options. 
- Second, authors need to be aware of the benefits of posting preprints and post-prints for constructive feedback [@bourne2017ten]. While it is clear preprints help establish precedence and allow findings to be openly accessible, it is still rare that constructive discussions form around preprints in an open form, which is consistent with findings on other preprint servers (e.g. bioRxiv [@anderson2020biorxiv]). Clearly, as a community, we need to provide better platforms that document discussions around preprint findings. Such discussions help the authors improve their work and communicate their findings more effectively (when done constructively, of course). One way to facilitate such discussions may be to use open peer reviewing services such as *Peer Community In* (PCI) or *PubPeer* to provide feedback on preprints that ultimately is associated with a preprint.
- Third, it remains clear that the time between preprinting and publication is still quite long. One possible explanation is that preprints are not being sent to suitable journals. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recommendations when publishing a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process. 
- Finally, while most preprints get seamlessly connected with and merged with their published version some remain disconnected. Incorrect cross-linking can create confusion and lead to frustration among authors. Thankfully, the reasons for unmatched preprints and publications are well understood and easily rectified. They often result from a mismatch between preprint and published metadata (e.g., titles and author details). In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%). We found that mismatched metadata almost always contributed to preprints and published articles not being matched automatically in Google Scholar. We recommend that authors ensure that, when preprints are published in a journal, they update their preprint with the publication DOI. Especially if their title has changed, this is quite easy for authors to do on *EcoEvoRxiv*, and it will ensure that the preprint is linked to the published article and that the citations are merged. It will also create a link between the final preprint and the published paper so that, no matter the open access status of the publication, the research findings will remain openly accessible, saving costs for authors and improving the visibility and use of research.

Despite the early successes of the new initiatives taken by *EcoEvoRxiv*, much work needs to be done to improve our community’s understanding and use of preprints (and post-prints). We view this perspective piece as one small step in achieving that goal. Authors must also be aware of the benefits of using community-driven preprint servers and the unique initiatives being pursued. Community-driven preprint servers are capable of setting their own agenda and are driven by the needs and desires of the community. Supporting these initiatives should be a priority for all researchers. Volunteers at *EcoEvoRxiv* should remain open to new and innovative ways to improve publication and open science practices. We believe that the future of preprints is bright, and community-driven initiatives, such as *EcoEvoRxiv*, are set to play a crucial role in the future of scientific publishing.

# Acknowledgements
We would like to thank the California Digital Library (CDL) and the CDL team (particularly Alainna Wrigley) for their support in hosting and maintaining *EcoEvoRxiv* for the Society for Open, Reliable, and Transparent Ecology and Evolution (SORTEE). We would also like to thank Gabriela Hidalgo and Daisy Larios for helping connect us with the IUCN and facilitating discussions to make *EcoEvoRxiv* a place where IUCN reports can be posted. Finally, we would also like to thank the endless number of SORTEE volunteers, and those especially on the *EcoEvoRxiv* Committee, who have helped to make *EcoEvoRxiv* a success. This paper emerged from a hackathon at the 2022 SORTEE conference, and we thank delegates who attended the session but could not be part of this paper. DWAN would also like to thank the Australian Research Council for a Future Fellowship (FT220100276).
